Last glacial cycle glacier erosion potential in the Alps

### 2021-05-27 articles/erosion/response-referee2.md

The field-based LGM outline was added on Fig. 2 (Alps-wide map of cumulative
erosion potential). In addition, the following sentences were added in the
"climate sensitivity" section of the discussion mentioning the general
eastwards bias of our results with regard to field evidence.

    It should be noted, however, that all runs presented here show a systematic
    bias with excessive glacier cover in the Eastern Alps and a diminished
    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
    et al., 2018). Thus the modelled patterns of erosion potential certainly
    includes a similar bias.

> Overall, the paper reports on a very worthwhile computer-intensive exercise.
> It is densely packed with results and repays a careful read and a study of
> the videos. Clearly there can be interesting comparisons with Lai and Anders
> (2021) very recent submission to esurf !

Indeed, this is quite a coincidence! A reference was added to the paper by Lai
and Anders (2021).

@@

> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
> map or hillshade of the Rhine area on the same scale should be provided.

In addition to the aforementioned changes to Fig. 4, a new panel was added for
the model final state at 0 ka, revealing the bedrock topography. So Fig. 4 now
has one panel during advance (36 ka), one during a maximum stage (24 ka), one
during retreat (16 ka), and one nearly ice-free (0 ka).

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

The figure was moved into the discussion of the "age of the glacial landscape"
to become Fig. 8, and is not longer referred to in the intro. The photo of
Lauterbrunnental was replaced by one of Bout du Monde in the Giffre Massif, and
the caption for the cirque photo now reads: "the unusually deep mountain cirque
revealed by the current demise of Chüebodengletscher" (I don't have a much
better photo for a high-elevation cirque, and glacial lakes from retreating
cirque glaciers are a common occurence in the Alps these days).

### 2021-05-27 articles/erosion/response-referee1.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Anonymous Referee #1

Dear Anonymous Referee #1,

Thank you very much for your positive comments and for your time spent
reviewing our manuscript. We try to address your comments one by one below, and
highlight relevant changes made to the manuscript.

@@

> The introduction gives a nice, relatively long (in relation to other
> chapters) overview on different aspects of glacial erosion and resulting
> features, and guides through problems to investigate them. Here, I don’t see
> very clear how different aspects or parts in the introduction are
> specifically addressed in this study, i.e. which problems are exactly aimed
> to be solved of this large portfolio of mentioned gaps in the understanding
> of glacial processes. I think this could be better formulated and balanced.

Our introduction is meant to revolve around two discussion points: the variable
imprint of glaciers on mountain topography, and the link between climate and
glacier erosion (somewhat adressed in the conclusions). This is followed by a
short methodological bibliography of process-based and empirical glacier
erosion laws, before we announce the scope of our work. A few key statements
were added to try and better tie these different sections.

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

Thank you for the suggestion. Figure 5 was amended with the
distribution of cumulative erosion potential over 100 m elevation bins.
The resulting plot shows that most modelled erosion occurs below 2000 m,
with the potential erosion volumes between 1000 and 2000 m elevation.

@@

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

Our calculations indeed produce locally high cumulative erosion potential in
the aforementioned crystalline massifs. More accurate modelling of small-scale
glaciers would be needed to provide an accurate answer to your question. Steep
topography, though, is
likely to be part of the explanation. A more substantial discussion of the high
erosion potential values in these areas and the model limitations that hinder
interpretation was added in the "age of the glacial landscape" subsection.
The new passage reads as follow:

@@

> Line 33 I think that’s not very easy to understand what landform you refer
> to? The “periglacial blockfields topped by glacial erratic boulders” (Wirsig
> et al., 2018)? Eventually be more precise on this would help readers.

Agreed. "Characteristic landform preservation" was replaced with "preserved
periglacial blockfields topped by erratic boulders".

> Line 101 Could be misleading as (glacial and periglacial) cirque erosion
> processes are not really covered by any glacial erosion law discussed here
> (or any other I am aware of, Sanders et al., 2012).

Thank you for pointing out this study. This particular sentenced was reworked
to acknowledge not only the limiting resolution but also the lack of relevant
erosion erosion processes. The reference was additionally incorporated in the
relevant part of the discussion (see general comments above).

@@

> Line 195 Very much share this view!

Good to know!

> Line 197 Would recommend to be more precise. Guess you know that there are
> many, many cirques in the Alps as low as 1500 m (and even below) e.g. forming
> in areas outside the connected ice stream network in the SW and easternmost
> Alps. This can be revealed by a quick check at any higher resolution DEM or
> google earth.

The statement was reworked and restricted to "the highest mountain cirques".

@@

> Valla, P.G., Shuster, D.L., van der Beek, P.A., 2011. Significant increase in
> relief of the European Alps during mid-Pleistocene glaciations. Nat. Geosci.
> 4, 688–692.

In addition to the aforementioned changes, the Sternai et al. (2013) reference
was added in the discussion of the vertical distribution of erosion with this
new sentence:

    Over longer timescales, though, the vertical distribution of erosion rates
    also depends on the erosional modification of topography (Sternai et al.,
    2013).

Thank you again for reading our work and giving us constructive feedback.

### 2021-05-27 articles/erosion/response-referee2.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Ian Evans

Dear Ian Evans,

Thank you very much for your and your colleague's time in preparing this
reviewing our manuscript. We try to address your comments one by one below, and
highlight relevant changes made to the manuscript.

@@

> The authors provide large amounts of very useful results from simulations,
> concerning the effects of varying glacier geometry on erosion potential. It
> was quite an achievement to compute this at fairly high temporal and spatial
> resolution (1 km and 50 a throughout the Alps). The conclusions are
> reasonable, and I value the results. They provide support for the
> traditional view, that high-altitude features such as cirques are eroded by
> steep, local glaciers and not under extensive icefields well above ELA. Most
> erosion is expected along major valleys (glacial troughs): the potential
> erosion is greatest low down the glaciers, with a lesser maximum at high
> altitude and a minimum between. The paper is well written and the Figures
> are all relevant and interesting. The qualitative conclusions are well
> justified.

> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
> sea level equivalent. I would love to see a map of the glaciers at that
> stage\*, which is well below maximum but considerably more than recent ice
> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).
>
>   [\*That stage can be approximated by pausing the first erosion
>    rates video around 99 ka and 13 ka. The three videos are strongly
>    recommended!]

Thank you for pointing this out. However, we are not confident that this result
is significant. The situation you describe occurs, for instance, during much of
MIS~5 (Fig. 2) and can be visualized in the animations (as you noted). During
such periods, glaciers are steep and only a few grid cells in width.

Increasing ice volume yields the build-up of less steep and thus slower-flowing
valley glaciers. Decreasing ice volume leads to (roughly) equally steep but
smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
laws), both yield a decrease of erosion volume, hence the local peak in erosion
volume. However, we would rather not include this as a result as the modelled
glacier velocities for such situations are unreliable due to the limiting
horizontal resolution and the shallow-ice physics. This latter fact was
highlighted in a new passage in the section on the "age of the glacial
landscape". Also including suggestions from reviewer #1, the new passage
reads as follow:

@@

Thank you. Figure 4 was reworked to include a map of Rhine glacier modelled
erosion rates at 36 ka, corresponding to the last major advance phase before
the LGM and with a similar extent to the deglacial stage of 16 ka. This also
corresponds to the topographic profiles shown on Fig. 8 and thus seems more
appropriate in hindsight.

TODO: also include complete maps as a supplement?

> I find no mention of lithology, rock type, geology, resistance or erodibility
> (excluding that citing Herman on line 56). This omission of half the erosion
> equation should be mentioned; it implies that the ‘potential’ nature of the
> erosion should be emphasised repeatedly. This adds to the admitted
> limitations of lack of feedback to changing topography, absence of hydrology
> and absence of subglacial sediment accumulation or movement. Using constant
> topography might be OK for the last glaciation, as here, but would become
> increasingly unrealistic for several glaciations. Destinations of eroded
> material are not considered: deposition in temporary storage may inhibit
> erosion in parts. The omission of hydrological processes must be a major
> drawback, as Herman et al. (2011), Egholm et al. (2012a, b) and Pedersen et
> al. (2014) showed that its inclusion of subglacial meltwater, plus horizontal
> stress gradients, gave more realistic simulations of erosion. For erosion
> rate as well as for cumulative erosion, it is potential erosion that is
> calculated, rather than predicted erosion.

The above limitations were made explicit with the following added sentences in
the methods subsection on the "erosion law":

@@

    Instead , we we assume that eroded material is instantly transported out of
    the system, thus neglegting its role in shielding the bedrock from glacier
    erosion in zones of temporary storage (Preusser et al., 2010). Neither do
    we account for differences in erosion effectiveness on different
    lithologies or erosion from subglacial and interglacial hydrologic
    processes. For the aforementioned reasons, we refer to the above computed
    rates, ė, as “potential erosion rates”.

Second and as implied, the entire text and the figure labels were reworked to
replace “erosion rates” with “potential erosion rates” nearly everywhere. To
avoid clutter a few sentences still mention “modelled erosion rates” instead of
“potential erosion rates”.

> A further necessary geographical qualification concerns the misfit between
> simulated and actual glacier extent. The latter is well established from
> mapping of moraines. It is necessary to check Seguinot et al. (2018) to see
> that all the models predict too much ice in the southwest (south of 45° N)
> and far too much in the east (Sava, Drava, Mur, Enns and Traun valleys),
> while the EPICA models produce too little in the west (Rhône and Jura). I
> agree that this is probably because while the amount of precipitation is
> changed, the spatial pattern is not: thus the orographic effect of ice
> build-up increasing precipitation shadow, and the circulation changes due to
> southward shift of westerlies at LGM, are not included. To make this paper
> more self-contained, a brief mention of the 2018 evaluation is in order.
> Once a more realistic LGM precipitation pattern is accepted, future modelling
> might use gradations between that and the present pattern as a function of
> temperature.

The field-based LGM outline was added on Fig. 2 (Alps-wide map of cumulative
erosion potential). In addition, the following sentences were added in the
"climate sensitivity" section of the discussion mentioning the general
eastwards bias of our results with regard to field evidence:

@@

Indeed. The sentence was corrected to include "advancing and retreating
glaciers" instead of "retreating glaciers".

## - Figures:

> In Fig. 3 is almost impossible to follow the brown lines, unless advancing
> and retreating parts of lines are distinguished (e.g. by colour). (The ice
> volume video may help out.)  Also: keep to ‘potential erosion volume’. The
> pale brown bars show a decline in annual erosion volume with ice volumes over
> 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more
> than ‘slight’.

Figure 3 was reworked to use different colours for periods of increasing and
decreasing ice volume. This actually strikingly highlights the two different
regimes, so thank you for the suggestion. The label was changed to 'potential
annual erosion volume'. In the text, 'slight' was replaced with a 'general'
tendency for slower erosion during periods of extensive glaciation.

@@

## Acknowledgements:

> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt
> Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.

> - Ian S. Evans, Durham University, U.K.

Thank you again to all of you for taking time to read our work and offer
constructive feedback.

### 2021-05-27 articles/erosion/alps-erosion.tex
Read through the responses once again.
Add reviewers in acknowledgements, tag as 2.2.

    \begin{figure*}
      \centerline{\includegraphics{alpero_cumulative}}
      \caption{%
        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
          potential over the last glacial cycle and geomorphological
          reconstruction of Last Glacial Maximum Alpine glacier extent for
          comparison \citep[solid red line,][]{Ehlers.etal.2011}.
          The background maps consists of the initial basal topography from
          SRTM \citep{Jarvis.etal.2008} and Natural Earth Data
          \citep{Patterson.Kelso.2017}.
        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
          equivalent (cm~s.l.e., black), annual (domain-integrated) potential
          erosion volume (light brown) and its 1-ka running mean (dark brown).
          Shaded gray areas indicate the timing for MIS~2 and~4
          \citep{Lisiecki.Raymo.2005}.
        \todo{hatches on all plots.}}
        \label{fig:cumulative}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of modelled rates in 100-m elevation bands
          across the entire model domain and its time evolution. Hatches
          indicate elevation bands with fewer than a hundred grid cells
          (100\,$km^2$).
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
        \todo{fix labels, test hatching below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}

@@

\begin{acknowledgements}
    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
    Aschwanden for their constant help and development with PISM, without whom
    this work would not have been possible. J.~Seguinot would like to thank
    John Jansen and Martin Margold for insightful early discussions, Susan
    Ivy-Ochs for being a constant (droid) motivator, Popi, Karsta and Stefan
    for welcoming him on Anafi and providing agreeable conditions for work in
    times of pandemic, and the Long Haul Trucker for conveying his laptop and
    data unbroken for 5000\,km across Europe and fourteen countries.
    I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and
    conversations while at University of Lausanne.
    The (previously published) ice flow simulations were made with computer
    resources provided by the Swiss National Supercomputing Centre (CSCS)
    allocations no.~s573 and sm13 to J.~Seguinot.
\end{acknowledgements}

>>>>

    \begin{figure*}
      \centerline{\includegraphics{alpero_cumulative}}
      \caption{%
        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
          potential over the last glacial cycle and geomorphological
          reconstruction of Last Glacial Maximum Alpine glacier extent for
          comparison \citep[solid red line,][]{Ehlers.etal.2011}.
          The background maps consists of the initial basal topography from
          SRTM \citep{Jarvis.etal.2008} and Natural Earth Data
          \citep{Patterson.Kelso.2017}.
        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
          equivalent (cm~s.l.e., black), annual (domain-integrated) potential
          erosion volume (light brown) and its 1-ka running mean (dark brown).
          Shaded gray areas indicate the timing for MIS~2 and~4
          \citep{Lisiecki.Raymo.2005}.}
        \label{fig:cumulative}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of modelled rates in 100-m elevation bands
          across the entire model domain and its time evolution. Hatches
          indicate elevation bands with fewer than a hundred grid cells
          (100\,$km^2$).
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.}
      \label{fig:hypsogram}

    \end{figure*}

@@

\begin{acknowledgements}
    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
    Aschwanden for their constant help and development with PISM, without whom
    this work would not have been possible. J.~Seguinot would like to thank
    John Jansen and Martin Margold for insightful early discussions, Susan
    Ivy-Ochs for being a constant (droid) motivator, Popi, Mattheos, Karsta,
    Stefan, Fran\c{c}oise and Dominique
    for welcoming him on Anafi and providing agreeable conditions for work in
    times of pandemic, and the Long Haul Trucker for conveying his laptop and
    data unbroken for 5000\,km across Europe and fourteen countries.
    I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and
    conversations while at University of Lausanne.
    An anonymous reviewer and Ian Evans (aided by Iestyn Barr, Jeremy Ely, Matt
    Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek) provided
    constructive feedback which has been incorporated in the present paper.
    The (previously published) ice flow simulations were made with computer
    resources provided by the Swiss National Supercomputing Centre (CSCS)
    allocations no.~s573 and sm13 to J.~Seguinot.
\end{acknowledgements}
